### 2015-05-03 writing/S'14 paper/join.tex
Caltrain edits.
Draft for arXiv.
Updates for arxiv
Sentence classification w/ TreeNNs

\section{Reasoning about semantic relations}\label{sec:join}

If a model is to learn the behavior of a relational logic like the one
presented here from a finite amount of data, it must learn to deduce new
relations from seen relations in a sound manner. The simplest such
deductions involve atomic statements using the relations in
Table~\ref{b-table}. For instance, given that $a \natrev b$ and $b
\natrev c$, one can conclude that $a \natrev c$, by basic
set-theoretic reasoning (transitivity of $\natrev$). Similarly, from
$a \natfor b$ and $b \natneg c$, it follows that $a \natalt c$.  The
full set of sound inferences of this form is summarized in
Table~\ref{tab:jointable}; cells containing a dot correspond to pairs
of relations for which no valid inference can be drawn in our logic.

% about the relations themselves that do not depend on the
% internal structure of the things being compared. For example, given
% that $a\sqsupset b$ and $b\sqsupset c$ one can conclude that
% $a\sqsupset c$ by the transitivity of $\sqsupset$, even without
% understanding $a$, $b$, or $c$. These seven relations support more
% than just transitivity: MacCartney and Manning's
% \cite{maccartney2009extended} join table defines 32 valid inferences
% that can be made on the basis of pairs of relations of the form $a R
% b$ and $b R' c$, including several less intuitive ones such as that if
% $a \natneg b$ and $b~|~c$ then $a \sqsupset c$.

\begin{table}[htp]
  \centering  
  \setlength{\arraycolsep}{8pt}
  \renewcommand{\arraystretch}{1.1}
  \newcommand{\UNK}{\cdot}  
  $\begin{array}[t]{c@{ \ }|*{7}{c}|}
    %\hline
    \multicolumn{1}{c}{}
             & \nateq     & \natfor     & \natrev     & \natneg    & \natalt     & \natcov     & \multicolumn{1}{c}{\natind} \\
    \cline{2-8}
    \nateq  & \nateq &   \natfor &  \natrev &  \natneg &   \natalt &  \natcov &  \natind \\
    \natfor & \natfor &  \natfor &  \UNK &  \natalt &   \natalt &  \UNK &  \UNK \\
    \natrev & \natrev &  \UNK &  \natrev &  \natcov &   \UNK &  \natcov &  \UNK \\
    \natneg & \natneg &  \natcov &  \natalt &  \nateq &    \natrev &  \natfor &  \natind \\
    \natalt & \natalt &  \UNK &  \natalt &  \natfor &   \UNK &  \natfor &  \UNK \\
    \natcov & \natcov &  \natcov &  \UNK &  \natrev &   \natrev &  \UNK &  \UNK \\
    \natind & \natind & \UNK &  \UNK &  \natind &  \UNK &  \UNK &  \UNK \\
    \cline{2-8}
  \end{array}$
  \caption{Inference path from premises $a\,R\,b$ (row) and $b\,S\,c$ (column) to the relation that holds between $a$ and $c$, if any.  These inferences are based on basic set-theoretic truths about the meanings of the underlying relations as described in Table~\ref{b-table}. We assess our model's ability to reproduce such inferential paths.}
  \label{tab:jointable}
\end{table}

@@

@@

    \caption{A train/test split of the atomic statements about the
      model.  Test statements not provable from the training data are
      crossed out.}
  \end{subfigure}  
  \caption{Small example structure and data for learning relation composition.}
  \label{lattice-figure}
\end{figure} 
% Note: None of these test examples is derivable from the shown training data, 
% but we suggest that there is additional training data, so we can cross lines
% out willy-nilly without fear.

% We test the model's ability to learn this behavior by creating
% artificial data sets of terms which represent sets of numbers. Since
% MacCartney and Manning's set of relations hold between sets as well as
% between sentences, we can use the underlying set structure to generate
% the all of the relations that hold between any pair of these terms, as
% in Figure \ref{lattice-figure}. We train the model defined above on a
% subset of these relations, but rather then presenting the model with a
% pair of tree-structured sentences as inputs, simply present it with
% two single terms, each of which corresponds to a single vector in the
% (randomly initialized) vocabulary matrix $V$, ensuring that the model
% has no information about the terms being compared except the relations
% between them.

We trained versions of both the RNN model and the RNTN model on these
data sets. In both cases, the models were implemented exactly as
described in Section~\ref{methods}, but since the items being compared
are single terms rather than full tree structures, the composition
layer was not involved, and the two models differed only in which
layer function was used for the comparison layer. We simply present
the models with two single terms, each of which corresponds to a
single vector in the (randomly initialized) vocabulary matrix $V$,
thereby ensuring that the model has no information about the terms
being compared except the relations between them.

\paragraph{Results} 
We found that the RNTN model worked best with 11-dimensional vector
representations for the 80 sets and a 90-dimensional feature vector
for the classifier, though the performance of the model was not highly
sensitive to either dimensionality parameter in any of the experiments
discussed here. This model was able to correctly label 99.4\% (94.3
macroaveraged F1) of the provably derivable test examples, and 95.4\%
(92.2 macro-F1) of the remaining test examples. The simpler RNN model worked
best with 11 and 75 dimensions, respectively, but was able to achieve
accuracies of only 89.0\% (80.7 macro-F1) and 62.5\% (40.7 macro-F1),
respectively. Training accuracy was 99.8\% for the RNTN and 95.4\% for
the RNN.

@@

% and create a dataset consisting of the relations between every pair of
% sets, yielding 6400 pairs. 3200 of these pairs are then chosen as a
% test dataset, and that test dataset was further split into the 2960
% examples that can be provably derived from the test data using
% MacCartney and Manning's join table (or by the symmetry of the
% relations in about half of the cases) and the 240 that
% cannot.

@@

These results are fairly straightforward to interpret. The RNTN model
was able to accurately encode the relations between the terms in the
geometric relations between their vectors, and was able to then use
that information to recover relations that were not overtly included
in the training data. In contrast, the RNN model was able to achieve
this behavior only incompletely. It is possible but not likely that it
could be made to find a good solution with further optimization on
different learning algorithms, or that it would do better on a larger
universe of sets, for which there would be a larger set of training
data to learn from, but the RNTN is readily able to achieve these
effects in the setting discussed here.

>>>>

\section{Reasoning about semantic labels}\label{sec:join}

If a model is to learn the behavior of a labelal logic like the one
presented here from a finite amount of data, it must learn to deduce new
labels from seen labels in a sound manner. The simplest such
deductions involve atomic statements using the labels in
Table~\ref{b-table}. For instance, given that $a \natrev b$ and $b
\natrev c$, one can conclude that $a \natrev c$, by basic
set-theoretic reasoning (transitivity of $\natrev$). Similarly, from
$a \natfor b$ and $b \natneg c$, it follows that $a \natalt c$.  The
full set of sound inferences of this form is summarized in
Table~\ref{tab:jointable}; cells containing a dot correspond to pairs
of labels for which no valid inference can be drawn in our logic.

% about the labels themselves that do not depend on the
% internal structure of the things being compared. For example, given
% that $a\sqsupset b$ and $b\sqsupset c$ one can conclude that
% $a\sqsupset c$ by the transitivity of $\sqsupset$, even without
% understanding $a$, $b$, or $c$. These seven labels support more
% than just transitivity: MacCartney and Manning's
% \cite{maccartney2009extended} join table defines 32 valid inferences
% that can be made on the basis of pairs of labels of the form $a R
% b$ and $b R' c$, including several less intuitive ones such as that if
% $a \natneg b$ and $b~|~c$ then $a \sqsupset c$.

\begin{table}[htp]
  \centering  
  \setlength{\arraycolsep}{8pt}
  \renewcommand{\arraystretch}{1.1}
  \newcommand{\UNK}{\cdot}  
  $\begin{array}[t]{c@{ \ }|*{7}{c}|}
    %\hline
    \multicolumn{1}{c}{}
             & \nateq     & \natfor     & \natrev     & \natneg    & \natalt     & \natcov     & \multicolumn{1}{c}{\natind} \\
    \cline{2-8}
    \nateq  & \nateq &   \natfor &  \natrev &  \natneg &   \natalt &  \natcov &  \natind \\
    \natfor & \natfor &  \natfor &  \UNK &  \natalt &   \natalt &  \UNK &  \UNK \\
    \natrev & \natrev &  \UNK &  \natrev &  \natcov &   \UNK &  \natcov &  \UNK \\
    \natneg & \natneg &  \natcov &  \natalt &  \nateq &    \natrev &  \natfor &  \natind \\
    \natalt & \natalt &  \UNK &  \natalt &  \natfor &   \UNK &  \natfor &  \UNK \\
    \natcov & \natcov &  \natcov &  \UNK &  \natrev &   \natrev &  \UNK &  \UNK \\
    \natind & \natind & \UNK &  \UNK &  \natind &  \UNK &  \UNK &  \UNK \\
    \cline{2-8}
  \end{array}$
  \caption{Inference path from premises $a\,R\,b$ (row) and $b\,S\,c$ (column) to the label that holds between $a$ and $c$, if any.  These inferences are based on basic set-theoretic truths about the meanings of the underlying labels as described in Table~\ref{b-table}. We assess our models' ability to reproduce such inferential paths.}
  \label{tab:jointable}
\end{table}

@@

% We test the model's ability to learn this behavior by creating
% artificial data sets of terms which represent sets of numbers. Since
% MacCartney and Manning's set of labels hold between sets as well as
% between sentences, we can use the underlying set structure to generate
% the all of the labels that hold between any pair of these terms, as
% in Figure \ref{lattice-figure}. We train the model defined above on a
% subset of these labels, but rather then presenting the model with a
% pair of tree-structured sentences as inputs, simply present it with
% two single terms, each of which corresponds to a single vector in the
% (randomly initialized) vocabulary matrix $V$, ensuring that the model
% has no information about the terms being compared except the labels
% between them.

We trained versions of both the RNN model and the RNTN model on these
data sets. In both cases, the models were implemented exactly as
described in Section~\ref{methods}, but since the items being compared
are single terms rather than full tree structures, the composition
layer was not involved, and the two models differed only in which
layer function was used for the comparison layer. We simply present
the models with two single terms, each of which corresponds to a
single vector in the (randomly initialized) vocabulary matrix $V$,
thereby ensuring that the model has no information about the terms
being compared except the labels between them.

\paragraph{Results} 
We found that the both models worked well with 15-dimensional vector
representations for the 80 set symbols,

@@

    \caption{A train/test split of the atomic statements about the
      model.  Test statements not provable from the training data are
      crossed out.}
  \end{subfigure}  
  \caption{Small example structure and data for learning label composition.}
  \label{lattice-figure}
\end{figure} 
% Note: None of these test examples is derivable from the shown training data, 
% but we suggest that there is additional training data, so we can cross lines
% out willy-nilly without fear.

@@

% and create a dataset consisting of the labels between every pair of
% sets, yielding 6400 pairs. 3200 of these pairs are then chosen as a
% test dataset, and that test dataset was further split into the 2960
% examples that can be provably derived from the test data using
% MacCartney and Manning's join table (or by the symmetry of the
% labels in about half of the cases) and the 240 that
% cannot.

@@

The RNTN model was able to accurately encode the labels between the terms in the
geometric labels between their vectors, and was able to then use
that information to recover labels that were not overtly included
in the training data. In contrast, the RNN model was able to achieve
this behavior only incompletely. It is possible but not likely that it
could be made to find a good solution with further optimization on
different learning algorithms, or that it would do better on a larger
universe of sets, for which there would be a larger set of training
data to learn from, but the RNTN is readily able to achieve these
effects in the setting discussed here.

### 2015-05-03 writing/S'14 paper/methods.tex
Sentence classification w/ TreeNNs

We use the model architecture proposed in \cite{bowman2013can} and
depicted in Figure~\ref{sample-figure}. The two phrases being compared
are built up separately on each side of the tree, using the same
composition function, until they have each been reduced to single
vectors. The resulting vectors are fed into a separate comparison
layer that is meant to generate a feature vector capturing the
relation between the two phrases. The output of this layer is then
given to a softmax classifier, which in turn produces a hypothesized
distribution over the seven relations represented in Table~\ref{b-table}.

For a composition layer, we evaluate models with both the plain neural
network layer function \eqref{rnn} and the RNTN layer function
\eqref{rntn} proposed in \citet{chen2013learning}. A sigmoid
nonlinearity (element-wise $\tanh$) is applied to the output of either
layer function, following \cite{socher2013acl1}.
%
\begin{gather} \label{rnn}
\vec{y}_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b}) \\ % TODO: Add column vectors?
\label{rntn}
\vec{y}_{\textit{RNTN}} = f(\vec{x}^{(l)T} \mathbf{T}^{[1 \ldots n]} \vec{x}^{(r)} + \mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b})
\end{gather} 
%
Here, $\vec{x}^{(l)}$ and $\vec{x}^{(r)}$ are the column vector
representations for the left and right children of the node, and
$\vec{y}$ is the node's output.  The RNN concatenates them, multiplies
them by an $n \times 2n$ matrix of learned weights, and applies the
element-wise non-linearity to the resulting vector. The RNTN has the
same basic structure, but with the addition of a learned third-order
tensor $\mathbf{T}$, dimension $n \times n \times n$, modeling
multiplicative interactions between the child vectors. Both models
include a bias vector, represented as $\vec{b}$.

>>>>

We use the model architecture proposed in \cite{bowman2013can} and
depicted in Figure~\ref{sample-figure}. The two phrases being compared
are built up separately on each side of the tree, using the same
composition function, until they have each been reduced to single
vectors. The resulting vectors are fed into a separate comparison
layer that is meant to generate a feature vector capturing the
label between the two phrases. The output of this layer is then
given to a softmax classifier, which in turn produces a hypothesized
distribution over the seven labels represented in Table~\ref{b-table}.

For a composition layer, we evaluate models with both the plain neural
network layer function \eqref{rnn} and the RNTN layer function
\eqref{rntn} proposed in \citet{chen2013learning}. A sigmoid
nonlinearity (element-wise $\tanh$) is applied to the output of either
layer function, following \cite{socher2013acl1}.
%
\begin{gather} \label{rnn}
\vec{y}_{\textit{RNN}} = f(\mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b}) \\ % TODO: Add column vectors?
\label{rntn}
\vec{y}_{\textit{RNTN}} = f(\vec{x}^{(l)T} \mathbf{T}^{[1 \ldots n]} \vec{x}^{(r)} + \mathbf{M} [\vec{x}^{(l)}; \vec{x}^{(r)}] + \vec{b})
\end{gather} 
%
Here, $\vec{x}^{(l)}$ and $\vec{x}^{(r)}$ are the column vector
representations for the left and right children of the node, and
$\vec{y}$ is the node's output.  The RNN concatenates them, multiplies
them by an $n \times 2n$ matrix of learned weights, and applies the
element-wise non-linearity to the resulting vector. The RNTN has the
same basic structure, but with the addition of a learned third-order
tensor $\mathbf{T}$, dimension $n \times n \times n$, modeling
multiplicative interactions between the child vectors. Both models
include a bias vector~$\vec{b}$.

### 2015-05-03 writing/S'14 paper/quantifiers.tex
Sentence classification w/ TreeNNs

\paragraph{Experiments}
Our experimental data consist of pairs of sentences generated from a
small artificial grammar. Each sentence contains a quantifier, a noun
which may be negated, and an intransitive verb which may be
negated. We use the basic quantifiers \ii{some}, \ii{most}, \ii{all},
\ii{two}, and \ii{three}, and their negations \ii{no}, \ii{not-all},
\ii{not-most}, \ii{less-than-two}, and \ii{less-than-three}. We also
include five nouns, four intransitive verbs, and the negation symbol
\ii{not}. In order to be able to define relations between sentences
with differing lexical items, we define the lexical relations between
each noun--noun pair, each verb--verb pair, and each
quantifier--quantifier pair. The grammar accepts aligned pairs of
sentences of this form and calculates the natural logic relationship
between them.  Some examples of these data are provided in
Table~\ref{examplesofdata}.  As in previous sections, the goal of
learning is then to assign these relational labels accurately to
unseen pairs of sentences.

@@

% To assign relation labels to sentence pairs, we built a small
% task-specific implemenation of MacCartney's logic that can
% accurately label sentences of this restricted language. The logic is
% not able to derive all intuitively true relations of this language,
% and fails to derive a single unique relation for certain types of
% statement, including De Morgean's laws (e.g. \ii{(all pets) growl
% $\natneg$ (some pet) (not growl)}), and we simply discard these
% examples. Exhaustively generating the valid sentences under this
% grammar and choosing those to which a relation label can be assigned
% yields 66k sentence pairs. Some examples of these data are provided
% in Table~\ref{examplesofdata}.

\begin{table}[htp]
  \centering
  \begin{tabular}{l c l}
    \toprule
    (most warthogs) walk         & $\natneg$ & (not-most warthogs) walk\\
    (most mammals) move          & $\natind$ &  (not-most (not turtles)) move\\
    (most (not pets)) (not swim) & $\natrev$ & (not-most (not pets)) move 
    \\[2ex]    
    (no turtles) (not growl)     & $\natalt$ & (no turtles) (not swim)\\
    (no warthogs) swim           & $\natrev$ & (no warthogs) move\\
    (no warthogs) move           & $\natfor$ & (no (not reptiles)) swim\\
    \bottomrule
  \end{tabular}
  \caption{Sample data involving two different quantifier pairs.}
  \label{examplesofdata}
\end{table}

We evaluate the model using two experimental settings. In the simpler
setting, \textsc{all split}, we randomly sample 85\% of the data and evaluate on the
remaining 15\%. In this setting, the model is being asked to learn a
complete reasoning system for the limited language and logic presented
in the training data, but it is not being asked to generalize to test
examples that are substantially different from those it was trained
on. Crucially, though, to succeed on this task, the model must be able
to recognize all of the lexical relations between the nouns, verbs,
and quantifiers and how they interact. For instance, it might see
\eqref{p1} and \eqref{p2} in training and, from that information,
determine \eqref{p3}.
%
% do not allow a blank line --- adds too much space
%
\begin{gather}
  \text{(most turtle) swim} \natalt \text{(no turtle) move}\label{p1}
  \\
  \text{(all lizard) reptile} \natfor  \text{(some lizard) animal}\label{p2}
  \\
  \text{(most turtle) reptile} \natalt \text{(no turtle) animal}\label{p3}
\end{gather}
%
% do not allow a blank line --- adds too much space
%
While our primary interest is in discovering the extent to which our
models can learn to encode the logic given an arbitrary amount of
data, we are also interested in the degree to which they can infer a
correct representation for the logic from more constrained training
data. To this end, we segment the sentence pairs according to which
quantifiers appear in each pair, and then hold out one such pair for
testing. We hypothesize that a model that can efficiently learn to
represent a logic should be able to construct an accurate
representation of each of the two quantifiers in the pair from the way that it
interacts with the other nine quantifiers with which it is seen in training. 
Since running this experiment requires choosing a pair of
quantifiers to hold out before training, the resource demands of
training prevent us from testing each of the 55 possible 
pairs of quantifiers, so we choose only four pairs to test on.  Three
of these (\ii{two}/\ii{less-than-two}, \ii{not-all}/\ii{not-most}, and
\ii{all}/\ii{some}) were chosen because they allow for the most
different class labels in the test data. The fourth is a
self-pair (\ii{no}/\ii{no}), meant to test that the model correctly
handles equality.

>>>>

\paragraph{Experiments}
Our experimental data consist of pairs of sentences generated from a
small artificial grammar. Each sentence contains a quantifier, a noun
which may be negated, and an intransitive verb which may be
negated. We use the basic quantifiers \ii{some}, \ii{most}, \ii{all},
\ii{two}, and \ii{three}, and their negations \ii{no}, \ii{not-all},
\ii{not-most}, \ii{less-than-two}, and \ii{less-than-three}. We also
include five nouns, four intransitive verbs, and the negation symbol
\ii{not}. In order to be able to define labels between sentences
with differing lexical items, we define the lexical labels between
each noun--noun pair, each verb--verb pair, and each
quantifier--quantifier pair. The grammar accepts aligned pairs of
sentences of this form and calculates the natural logic labelship
between them.  Some examples of these data are provided in
Table~\ref{examplesofdata}.  As in previous sections, the goal of
learning is then to assign these labelal labels accurately to
unseen pairs of sentences.

@@

% To assign label labels to sentence pairs, we built a small
% task-specific implemenation of MacCartney's logic that can
% accurately label sentences of this restricted language. The logic is
% not able to derive all intuitively true labels of this language,
% and fails to derive a single unique label for certain types of
% statement, including De Morgean's laws (e.g. \ii{(all pets) growl
% $\natneg$ (some pet) (not growl)}), and we simply discard these
% examples. Exhaustively generating the valid sentences under this
% grammar and choosing those to which a label label can be assigned
% yields 66k sentence pairs. Some examples of these data are provided
% in Table~\ref{examplesofdata}.

\begin{table}[htp]
  \centering
  \begin{tabular}{l c l}
    \toprule
    (most warthogs) walk         & $\natneg$ & (not-most warthogs) walk\\
    (most mammals) move          & $\natind$ &  (not-most (not turtles)) move\\
    (most (not pets)) (not swim) & $\natrev$ & (not-most (not pets)) move\\ 
    %\\[2ex]    
    %(no turtles) (not growl)     & $\natalt$ & (no turtles) (not swim)\\
    %(no warthogs) swim           & $\natrev$ & (no warthogs) move\\
    %(no warthogs) move           & $\natfor$ & (no (not reptiles)) swim\\
    \bottomrule
  \end{tabular}
  \caption{Sample data involving two different quantifier pairs.}
  \label{examplesofdata}
\end{table}

We evaluate the model using two experimental settings. In the simpler
setting, \textsc{all split}, we randomly sample 85\% of the data and evaluate on the
remaining 15\%. In this setting, the model is being asked to learn a
complete reasoning system for the limited language and logic presented
in the training data, but it is not being asked to generalize to test
examples that are substantially different from those it was trained
on. Crucially, though, to succeed on this task, the model must be able
to recognize all of the lexical labels between the nouns, verbs,
and quantifiers and how they interact. For instance, it might see
\eqref{p1} and \eqref{p2} in training and, from that information,
determine \eqref{p3}.
%
% do not allow a blank line --- adds too much space
%
\begin{gather}
  \text{(most turtle) swim} \natalt \text{(no turtle) move}\label{p1}
  \\
  \text{(all lizard) reptile} \natfor  \text{(some lizard) animal}\label{p2}
  \\
  \text{(most turtle) reptile} \natalt \text{(no turtle) animal}\label{p3}
\end{gather}
%
% do not allow a blank line --- adds too much space
%
While our primary interest is in discovering the extent to which our
models can learn to encode the logic given an arbitrary amount of
data, we are also interested in the degree to which they can infer a
correct representation for the logic from more constrained training
data. To this end, we segment the sentence pairs according to which
quantifiers appear in each pair, and then hold out one such pair for
testing. We hypothesize that a model that can efficiently learn to
represent a logic should be able to construct an accurate
representation of each of the two quantifiers in the pair from the way that it
interacts with the other nine quantifiers with which it is seen in training. 
Since running this experiment requires choosing a pair of
quantifiers to hold out before training, the resource demands of
training prevent us from testing each of the 55 possible 
pairs of quantifiers, so we choose only four pairs to test on.  Three
of these (\ii{two}/\ii{less-than-two}, \ii{not-all}/\ii{not-most}, and
\ii{all}/\ii{some}) were chosen because they allow for the most
different class labels in the test data. The fourth is a
self-pair (\ii{no}/\ii{no}), meant to test that the model correctly
handles equality.
